---
title: "Jeffrey_hw8_exercise1"
author: "Jack Jeffrey"
format: pdf
editor: visual
---

## HW8

```{r}
# Set CRAN mirror to avoid errors during package installation
options(repos = c(CRAN = "https://cran.rstudio.com/"))
# Set working directory and load world data
setwd("/Users/jackjeffrey/Documents/Poli502_Jeffrey/Data")
world <- read.csv("world.csv")
# Investigate dataset
head(world)
str(world)
names(world)
```
# 1. Estimate a constant-only model
```{r}
# Fit a constant only model
model_constant <- lm(women09 ~ 1, data = world)
# View summary of model
summary(model_constant)
# Estumate of intercept/mean equals 17.17772 which is 17% female representation
# across all countries on average, standard error of 0.8238, P value below zero
# indicates the result is statistically significant meaning there is a clear average.
# 11O Observations not included due to NA's.
```
# 2. Estimate a model that uses per capita GDP (gdp_10_thou) as the main independent
# variable.
```{r}
# Fitting a linear model with women09 variable is dependent variable and gdp_10_thou
# as independent variable 
model_gdp <- lm(women09 ~ gdp_10_thou, data = world)
# view summary of model
summary(model_gdp)
#  The intercept is 14.8430. When GDP (gdp_10_thou) is 0, the baseline female
# representation would be approximately 14.84%.The coefficient for GDP is 3.4574, #bfor each increase of 1,000 dollars in per capita GDP, female representation is
# expected to increase by 3.46%. The p-value for gdp_10_thou is 5.5e-05, which is
# much smaller than 0.05. This means that the relationship between GDP and female
# representation is statistically significant, and we 
# can reject the null hypothesis that there is no relationship. 
# The R-squared value is low (9.3%), which suggests that while GDP 
# is statistically significant, it explains only a small portion 
# of the variation in female representation.
```
# 3. Create a graph that shows the estimated effect of per capita
# GDP on female representation using the effect function.

```{r}
# Istall effects package 
install.packages("effects")
library(effects)
# Create effect graph for gdp per capitas effect on female representation
effect_gdp <- effect("gdp_10_thou", model_gdp)
plot(effect_gdp, main = "Estimated Effect of Per Capita GDP on Female Representation",
     xlab = "Per Capita GDP (in 10,000 dollars)", ylab = "Female Representation (Women09)")
```
# 4. Estimate a model that uses a dummy variable that measures
# electoral system (pr_sys) as the main independent variable.

```{r}
# Investigate PR system variable to see if its is binary
unique(world$pr_sys)
# Estimate a model with PR system as the main independent variable
model_pr_sys <- lm(women09 ~ pr_sys, data = world)
# summarize result
summary(model_pr_sys)
# Proportional representation has a positive effect on female representation,
# with a coefficient of 8.23. Pvalue is 7.02e-07 and highly significant, 
# meaning the electoral system has a statistically significant impact 
# on female representation. The R-squared value is quite low, suggesting 
# that the electoral system alone explains only about 12.95% of the 
# variation in female representation. For countries with a non-proportional
# representation electoral system, the average 
# female representation is about 14.16.
```
# 5. Create a graph that shows the estimated effect of electoral
# system on female representation using the effect function.
```{r}
# Create an effect graph for the effect of electoral systems 
# on female representation
effect_pr_sys <- effect("pr_sys", model_pr_sys)
# Plot the effect 
plot(effect_pr_sys)
# Based on the plot proportional representation systems result in 
# much higher female representation 
```
# 6. Estimate a model that includes per capita GDP AND electoral
# system dummy at the same time.
```{r}
# Create a multiple regression model
model_gdp_prsys <- lm(women09 ~ gdp_10_thou + pr_sys, data = world)
# View results
summary(model_gdp_prsys)
# Interpretation - The residual standard error of 9.73 is the average difference
# between the actual values of female representation and the values 
# predicted by the model. 20.77%  of the variation in 
# female representation is explained by the per capita GDP 
# and electoral system combined based on the R squared value. 
# With a p-value of 4.049e-09, this model is highly significant.
```
# 7. Comparing the four models you have estimated so far, which
# one fit the data best?

```{r}
# Model 4 fits the data best of the four models as it has 
# the lowes Residual standard error of the four models at 9.73. It also
# has the highest adjusted R squared meaning it accounts for more of 
# the variation in female representation after penalizing for 
# added predictors
```

# 8. Create a graph that shows the effect of per capita GDP
# on Y for countries that adopt proportional representation system.

```{r}
# filter out missing data in pr sys
pr_data <- subset(world, pr_sysYes == 1)

# Define the model
pr_model <- lm(women09 ~ gdp_10_thou, data = pr_data)

# Generate predictions 
gdp_seq <- seq(min(pr_data$gdp_10_thou, na.rm = TRUE), 
               max(pr_data$gdp_10_thou, na.rm = TRUE), 
               length.out = 100)

# Create data frame from predictions 
prediction_data <- data.frame(gdp_10_thou = gdp_seq)

# Predict Y for GDP values
prediction_data$women09_pred <- predict(pr_model, newdata = prediction_data)

# Load tidyverse 
library(tidyverse)

# Graph plot 
ggplot(pr_data, aes(x = gdp_10_thou, y = women09)) +
  geom_point(alpha = 0.5) +  # Scatterplot of actual data
  geom_line(data = prediction_data, aes(x = gdp_10_thou, y = women09_pred), 
            color = "blue", size = 1) +  # Predicted effect line
  labs(
    title = "Effect of Per Capita GDP on Female Representation",
    x = "Per Capita GDP (in thousands)",
    y = "Female Representation (women09)"
  ) +
  theme_minimal()
```
# 9. Try creating the same graph by providing "gdp_10_thou:pr_sys" as the term.

```{r}
# Model with interaction term
model_interaction <- lm(women09 ~ gdp_10_thou * pr_sys, data = world)

# Plot effect for interaction term
effect_gdp_pr_interaction <- Effect(
  focal.predictors = c("gdp_10_thou", "pr_sys"), 
  mod = model_interaction, 
  given.values = list(pr_sysYes = 1)
)

# Plot graph
plot(effect_gdp_pr_interaction)

```
# 10. Estimate a regression model of female representation that uses 
# region as the main independent variable. 

```{r}
# Ensure region is a factor variable 
world$region <- as.factor(world$region)
# Estimate regression model 
model_region <- lm(women09 ~ region, data = world)
# View model summary 
summary(model_region)
# the regions with statistically significant results are -  Asia-Pacific, Middle
# East, Scandinavia, and Western Europe.
# The Scandinavia region has the highest positive impact 
# on female representation, while the Middle East has the 
# largest negative impact. The Multiple R-squared of 0.2615 
# indicates that approximately 26.15% of the variability 
# in female representation is explained by the region variable.
```
# 11. Create an effect plot that shows the relationship between region and 
# female representation. 

```{r}
# Fit the model with refion as the main independent variable 
model_region <- lm(women09 ~ region, data = world)
# Generate effect of region on female representation 
effect_region <- Effect(focal.predictors = "region", mod = model_region)
# Plot the effect 
plot(effect_region, main = "Effect of Region on Female Representation",
     xlab = "Region", ylab = "Female Representation (Women09)", 
     col = "blue", cex.lab = 1.2, cex.axis = 1.2)
```
# 12. Estimate a regression model of female representation on per cpaita
# GDP that controls for region. Based on the results, can we say 
# per capita GDP is an important determinant of female representation?
# Why or why not?

```{r}
# Fit the model with per capita gdp and region as independent variables 
model_gdp_region <- lm(women09 ~ gdp_10_thou + region, data = world)
# View model summary 
summary(model_gdp_region)
# Per capita GDP is not an important determinant of female representation 
# in this model. The coefficient for gdp_10_thou is 
# not statistically significant (p-value = 0.3917). 
# This suggests that, when controlling for region, 
# changes in per capita GDP do not have a strong 
# or statistically significant impact on female 
# representation across countries. 
# The p-value for gdp_10_thou is greater than 0.05, 
# which means there is not enough evidence to reject the null. 
```
# 13. Estimate a regression model of female representation on frac_eth3, a 
# three-category ordinal variable that measures levels of ethnic fractionalization.

```{r}
# Create a model using ethnic fractionalization as a categorical nominal variable
model_frac_eth3 <- lm(women09 ~ factor(frac_eth3), data = world)
# view summary of model
summary(model_frac_eth3)
```
# 14. Based on the results, do you think ethnic fractionalization levels
# have a positive/negative impact on female representation? 
```{r}
# Based on the regression results, 
# we cannot conclude that ethnic fractionalization 
# levels have a statistically significant positive 
# or negative impact on female
# representation in this dataset. 
# The p-values for all categories of ethnic fractionalization 
# are high, and the model as a whole is not significant.
```


