---
title: "Jeffrey_hw8"
author: "Jack Jeffrey"
format: pdf
editor: visual
---

## HW8
# setup
```{r}

# Set consistent root directory for all chunks
knitr::opts_knit$set(root.dir = "/Users/jackjeffrey/Documents/Poli502_Jeffrey/Data")
# Set CRAN mirror to avoid errors during package installation
options(repos = c(CRAN = "https://cran.rstudio.com/"))

install.packages("effects")  # Only if not already installed
library(effects)


getwd()


setwd("/Users/jackjeffrey/Documents/Poli502_Jeffrey/Data")


world <- read.csv("world.csv")


world <- read.csv("/Users/jackjeffrey/Documents/Poli502_Jeffrey/Data/world.csv")


```

# 1. Estimate a constant-only model

```{r}
# Fit a constant only model
model_constant <- lm(women09 ~ 1, data = world)
# View summary of model
summary(model_constant)
# Estimate of intercept/mean equals 17.17772 which is 17% female representation
# across all countries on average, standard error of 0.8238, P-value below zero
# indicates the result is statistically significant meaning there is a clear 
# average.
# 110 Observations not included due to NA's.
```

# 2. Estimate a model that uses per capita GDP (gdp_10_thou) as the main independent variable.

```{r}
# Fitting a linear model with women09 variable as dependent and gdp_10_thou
# as independent variable
model_gdp <- lm(women09 ~ gdp_10_thou, data = world)
# view summary of model
summary(model_gdp)
# The intercept is 14.8430. When GDP (gdp_10_thou) is 0, the baseline female
# representation would be approximately 14.84%. The coefficient for GDP is 3.4574, meaning
# for each increase of 1,000 dollars in per capita GDP, female representation is
# expected to increase by 3.46%. The p-value for gdp_10_thou is 5.5e-05, which is
# much smaller than 0.05. This means that the relationship between GDP and female
# representation is statistically significant.
# The R-squared value is low (9.3%), which suggests that while GDP 
# is statistically significant, it explains only a small portion 
# of the variation in female representation.
```

# 3. Create a graph that shows the estimated effect of per capita GDP on female representation

# using the effect function.

```{r}
# Create effect graph for GDP per capita's effect on female representation
effect_gdp <- effect("gdp_10_thou", model_gdp)
plot(effect_gdp, main = "Estimated Effect of Per Capita GDP on Female Representation",
     xlab = "Per Capita GDP (in 10,000 dollars)", ylab = "Female Representation (Women09)")
```

# 4. Estimate a model that uses a dummy variable that measures electoral system (pr_sys) as the

# main independent variable.

```{r}
# Investigate PR system variable to see if it is binary
unique(world$pr_sys)
# Estimate a model with PR system as the main independent variable
model_pr_sys <- lm(women09 ~ pr_sys, data = world)
# summarize result
summary(model_pr_sys)
# Proportional representation has a positive effect on female representation,
# with a coefficient of 8.23. P-value is 7.02e-07 and highly significant, 
# meaning the electoral system has a statistically significant impact 
# on female representation. The R-squared value is quite low, suggesting 
# that the electoral system alone explains only about 12.95% of the 
# variation in female representation. For countries with a non-proportional
# representation electoral system, the average 
# female representation is about 14.16.
```

# 5. Create a graph that shows the estimated effect of electoral system on female representation

# using the effect function.

```{r}
# Create an effect graph for the effect of electoral systems 
# on female representation
effect_pr_sys <- effect("pr_sys", model_pr_sys)
# Plot the effect 
plot(effect_pr_sys)
# Based on the plot, proportional representation systems result in 
# much higher female representation
```

# 6. Estimate a model that includes per capita GDP AND electoral system dummy at the same time.

```{r}
# Create a multiple regression model
model_gdp_prsys <- lm(women09 ~ gdp_10_thou + pr_sys, data = world)
# View results
summary(model_gdp_prsys)
# Interpretation - The residual standard error of 9.73 is the average difference
# between the actual values of female representation and the values 
# predicted by the model. 20.77%  of the variation in female representation 
# is explained by the per capita GDP and electoral system combined based on the R-squared value. 
# With a p-value of 4.049e-09, this model is highly significant.
```

# 7. Comparing the four models you have estimated so far, which one fits the data best?

```{r}
# Model 4 fits the data best of the four models as it has 
# the lowest Residual Standard Error and highest Adjusted R-squared 
# meaning it accounts for more of the variation in female representation after penalizing for added predictors
```

# 8. Create a graph that shows the effect of per capita GDP on Y for countries that adopt a

# proportional representation system.

```{r}
# filter out missing data in pr_sys (ensure pr_sys is a factor with 1/0 values or "Yes"/"No")
world$pr_sys <- as.factor(world$pr_sys)  # make sure it's a factor
levels(world$pr_sys)  # check factor levels to ensure correct filtering

# Filter data for countries with proportional representation system
pr_data <- subset(world, pr_sys == "Yes")  # Adjust to match your factor level

# Define the model
pr_model <- lm(women09 ~ gdp_10_thou, data = pr_data)

# Generate predictions
gdp_seq <- seq(min(pr_data$gdp_10_thou, na.rm = TRUE), 
               max(pr_data$gdp_10_thou, na.rm = TRUE), 
               length.out = 100)

# Create data frame from predictions
prediction_data <- data.frame(gdp_10_thou = gdp_seq)

# Predict Y for GDP values
prediction_data$women09_pred <- predict(pr_model, newdata = prediction_data)

# Load tidyverse
library(tidyverse)

# Graph plot
ggplot(pr_data, aes(x = gdp_10_thou, y = women09)) +
  geom_point(alpha = 0.5) +  # Scatterplot of actual data
  geom_line(data = prediction_data, aes(x = gdp_10_thou, y = women09_pred), 
            color = "blue", size = 1) +  # Predicted effect line
  labs(
    title = "Effect of Per Capita GDP on Female Representation",
    x = "Per Capita GDP (in thousands)",
    y = "Female Representation (women09)"
  ) +
  theme_minimal()
```

# 9. Try creating the same graph by providing "gdp_10_thou:pr_sys" as the term.

```{r}
# Model with interaction term
model_interaction <- lm(women09 ~ gdp_10_thou * pr_sys, data = world)

# Plot effect for interaction term
effect_gdp_pr_interaction <- Effect(
  focal.predictors = c("gdp_10_thou", "pr_sys"), 
  mod = model_interaction, 
  given.values = list(pr_sysYes = 1)
)

# Plot graph
plot(effect_gdp_pr_interaction)

```

# 10. Estimate a regression model of female representation that uses region as the main

# independent variable.

```{r}
# Ensure region is a factor variable
world$region <- as.factor(world$region)
# Estimate regression model
model_region <- lm(women09 ~ region, data = world)
# View model summary
summary(model_region)
# The regions with statistically significant results are - Asia-Pacific, Middle
# East, Scandinavia, and Western Europe.
# The Scandinavia region has the highest positive impact 
# on female representation, while the Middle East has the 
# largest negative impact. The Multiple R-squared of 0.2615 
# indicates that approximately 26.15% of the variability 
# in female representation is explained by the region variable.
```

# 11. Create an effect plot that shows the relationship between region and female

# representation.

```{r}
# Generate effect of region on female representation
effect_region <- Effect(focal.predictors = "region", mod = model_region)
# Plot the effect
plot(effect_region, main = "Effect of Region on Female Representation",
     xlab = "Region", ylab = "Female Representation (Women09)", 
     col = "blue", cex.lab = 1.2, cex.axis = 1.2)
```

# 12. Estimate a regression model of female representation on per capita GDP that controls for

# region.

```{r}
# Fit the model with per capita GDP and region as independent variables
model_gdp_region <- lm(women09 ~ gdp_10_thou + region, data = world)
# View model summary
summary(model_gdp_region)
# The model shows the effect of GDP after controlling for region.
# The Adjusted R-squared of 0.2643 is slightly higher than previous models.
```

# \# 13. Estimate a regression model of female representation on frac_eth3, a

# three-category ordinal variable that measures levels of ethnic fractionalization.

```{r}
# Create a model using ethnic fractionalization as a categorical nominal variable
model_frac_eth3 <- lm(women09 ~ factor(frac_eth3), data = world)
# view summary of model
summary(model_frac_eth3)
```

# 14. Based on the results, do you think ethnic fractionalization levels

# have a positive/negative impact on female representation?

```{r}
# Based on the regression results, 
# we cannot conclude that ethnic fractionalization 
# levels have a statistically significant positive 
# or negative impact on female
# representation in this dataset. 
# The p-values for all categories of ethnic fractionalization 
# are high, and the model as a whole is not significant.
```
